---
title: "Reference library coverage of Thames fish taxa"
author: "Wanda Bodnar"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

# Overview

This document assesses the coverage of Thames fish taxa recorded in Environment Agency (EA) surveys within the [Meta-Fish-Lib reference database](https://github.com/genner-lab/meta-fish-lib) for the MiFish and Tele02 primer sets.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ape)
library(patchwork)
```

# Load reference libraries

```{r load-reference-libraries}
source("https://raw.githubusercontent.com/genner-lab/meta-fish-lib/main/scripts/references-load-remote.R")
source("https://raw.githubusercontent.com/genner-lab/meta-fish-lib/main/scripts/references-clean.R")

mifish_lib <- subset_references(
  df = reflib.cleaned,
  metabarcode = "12s.miya" ) %>%
  derep_filter(derep = TRUE, proplen = 0.5) %>%
  distinct(sciNameValid)

tele02_lib <- subset_references(
  df = reflib.cleaned,
  metabarcode = "12s.taberlet") %>%
  derep_filter(derep = TRUE, proplen = 0.5) %>%
  distinct(sciNameValid)
```

# Load and clean Environment Agency data

```{r load-clean-data}
ea_data <- read_csv("TRAC_Fish_Counts.csv")

ea_species <- ea_data %>%
  mutate(LATIN_NAME = str_trim(LATIN_NAME)) %>%
  filter(!is.na(LATIN_NAME), LATIN_NAME != "") %>%
  distinct(LATIN_NAME) %>%
  filter(!str_detect(LATIN_NAME, "idae$")) %>%
  filter(!str_detect(LATIN_NAME, " sp\\.")) %>%
  filter(!str_detect(LATIN_NAME, " x | × "))
```

# Compare reference library coverage

```{r comparison-matrix}
comparison_matrix <- ea_species %>%
  mutate(
    In_MiFish = LATIN_NAME %in% mifish_lib$sciNameValid,
    In_Tele02 = LATIN_NAME %in% tele02_lib$sciNameValid,
    Category = case_when(
      In_MiFish & In_Tele02  ~ "In both databases",
      In_MiFish & !In_Tele02 ~ "Only in MiFish",
      !In_MiFish & In_Tele02 ~ "Only in Tele02",
      TRUE                   ~ "Missing from both"))
```

# Summary statistics

```{r summary-statistics}
coverage_summary <- comparison_matrix %>%
  summarise(
    Total_species = n(),
    MiFish = sum(In_MiFish),
    Tele02 = sum(In_Tele02))

coverage_summary
```

# Coverage pie chart

```{r coverage-pie-chart, fig.width=7, fig.height=6}
coverage_cols <- c(
  "In both databases" = "#2ecc71",
  "Only in MiFish" = "#f39c12",
  "Only in Tele02" = "#3498db",
  "Missing from both" = "#e74c3c")

pie_data <- comparison_matrix %>%
  count(Category) %>%
  mutate(Percentage = n / sum(n))

plot_coverage <- ggplot(
  pie_data,
  aes(x = "", y = Percentage, fill = Category)) +
  geom_col(colour = "white", width = 1) +
  coord_polar("y") +
  scale_fill_manual(values = coverage_cols) +
  geom_text(
    aes(label = scales::percent(Percentage, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    size = 4
  ) +
  theme_void() +
  labs(title = "Reference library coverage of Thames fish taxa") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom",
    legend.title = element_blank())

plot_coverage
```

# Missing taxa

```{r missing-taxa}
missing_species <- comparison_matrix %>%
  filter(!In_MiFish | !In_Tele02) %>%
  arrange(Category, LATIN_NAME)

missing_species
```

# Library coverage matrix

```{r coverage-matrix, fig.width=7, fig.height=10}
tile_data <- comparison_matrix %>%
  mutate(
    Sort_Order = case_when(
      Category == "Missing from both" ~ 1,
      Category %in% c("Only in MiFish", "Only in Tele02") ~ 2,
      TRUE ~ 3)) %>%
  arrange(Sort_Order, LATIN_NAME) %>%
  mutate(
    LATIN_NAME = factor(
      LATIN_NAME,
      levels = rev(unique(LATIN_NAME)))) %>%
  pivot_longer(
    cols = c(In_MiFish, In_Tele02),
    names_to = "Library",
    values_to = "Present") %>%
  mutate(
    Library = recode(
      Library,
      In_MiFish = "MiFish",
      In_Tele02 = "Tele02"),
    Status = if_else(
      Present,
      "Sequence available",
      "Missing from library"))

plot_matrix <- ggplot(
  tile_data,
  aes(x = Library, y = LATIN_NAME)) +
  geom_tile(
    aes(fill = Status),
    colour = "white",
    linewidth = 0.3) +
  scale_fill_manual(
    values = c(
      "Sequence available" = "#2ecc71",
      "Missing from library" = "#e74c3c")) +
  labs(
    title = "Reference library coverage for Thames fish taxa",
    subtitle = "Species missing from both libraries are shown first",
    x = NULL,
    y = NULL,
    fill = NULL) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey40"),
    axis.text.y = element_text(
      face = "italic",
      size = 7),
    panel.grid = element_blank(),
    legend.position = "top")

plot_matrix
```

# Abundance-weighted coverage

```{r abundance-weighted-coverage}
species_abundance <- ea_data %>%
  group_by(LATIN_NAME) %>%
  summarise(
    Total_Fish = sum(FISH_COUNT, na.rm = TRUE),
    .groups = "drop")

coverage_weighted <- comparison_matrix %>%
  left_join(species_abundance, by = "LATIN_NAME")

species_coverage <- mean(
  comparison_matrix$In_MiFish |
    comparison_matrix$In_Tele02) * 100

abundance_coverage <- 100 * (
  sum(
    coverage_weighted$Total_Fish[
      coverage_weighted$In_MiFish |
        coverage_weighted$In_Tele02]) /
    sum(coverage_weighted$Total_Fish))
```

# Species versus abundance coverage

```{r weighted-coverage-plot, fig.width=7, fig.height=5}
coverage_plot_data <- tibble(
  Metric = c(
    "Species coverage",
    "Abundance-weighted coverage"),
  Coverage = c(
    species_coverage,
    abundance_coverage))

plot_weighted <- ggplot(
  coverage_plot_data,
  aes(x = Metric, y = Coverage, fill = Metric)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(round(Coverage, 1), "%")),
    vjust = -0.5,
    size = 5) +
  scale_y_continuous(
    limits = c(0, 105),
    expand = expansion(mult = c(0, 0))) +
  scale_fill_manual(
    values = c(
      "Species coverage" = "#3498db",
      "Abundance-weighted coverage" = "#2ecc71")) +
  labs(
    title = "Reference library completeness of Thames fish taxa recorded in EA surveys",
    subtitle = "Assessment using Meta-Fish-Lib reference sequences for MiFish and Tele02",
    x = NULL,
    y = "Coverage (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none")

plot_weighted
```